Finally Tagless, Partially Evaluated

### 2008-02-26 tagless/tagless-final.tex
Added ref and discussion of Lars Berkedal's work.

Intuitively, self\hyp interpretation is straightforward in our
framework: the functions comprising the interpreters
in~\S\ref{S:interpreter-RL} may just as well be written in our object
language:~\Fig.\ref{fig:self-eval}.
\begin{figure}
\begin{align*}
    \ident{int} &= \fun{x} x \\
    \ident{add} &= \fun{x} \fun{y} x+y \\
    \ident{if\_}&= \fun{b} \fun{t} \fun{e} \cond{b}{t\,0}{e\,0}
        \displaybreak[0] \\
    \ident{lam} &= \fun{f} f \\
    \ident{app} &= \fun{f} \fun{x} fx \\
    \ident{fix} &= \fun{g} \fix{f} \fun{x} gfx
\end{align*}
\caption{The \emph{object} functions implementing an
evaluator.  (We use the number~$0$ in lieu of a unit value.)}
\label{fig:self-eval}
\end{figure}
We thus map each object term~$e$ to an object term~$\encode{e}$ as follows.
We call this mapping \emph{pre-encoding}.
\begin{align*}
    \encode{x} &= x \\
    \encode{n} &= \ident{int}\, n \\
    \encode{e_1 + e_2} &= \ident{add} \encode{e_1} \encode{e_2} \\
    \encode{\cond{b}{t}{e}} &= \rlap{$\ident{if\_} \encode{b}
        \left(\fun{\_}\encode{t}\right) \left(\fun{\_}\encode{e}\right)$}
        \displaybreak[0] \\
    \encode{\fun{x}e} &= \ident{lam} (\fun{x} \encode{e}) \\
    \encode{fx} &= \ident{app} \encode{f} \encode{x} \\
    \encode{\fix{f}e} &= \ident{fix} (\fun{f} \encode{e})
\end{align*}
The metavariables $x$ and~$n$ stand for a variable and an integer,
respectively.
This pre-encoding is just like how we represent object terms in the
metalanguage in the preceding sections, but it produces
terms in the object language rather than the metalanguage.

@@

>>>>

Intuitively, self\hyp interpretation is straightforward in our
framework: the functions comprising the interpreters
in~\S\ref{S:interpreter-RL} may just as well be written in our object
language:~Fig.~\ref{fig:self-eval}.
\begin{figure}
\begin{align*}
    \ident{int} &= \fun{x} x \\
    \ident{add} &= \fun{x} \fun{y} x+y \\
    \ident{if\_}&= \fun{b} \fun{t} \fun{e} \cond{b}{t\,0}{e\,0}
        \displaybreak[0] \\
    \ident{lam} &= \fun{f} f \\
    \ident{app} &= \fun{f} \fun{x} fx \\
    \ident{fix} &= \fun{g} \fix{f} \fun{x} gfx
\end{align*}
\caption{The \emph{object} functions implementing an
evaluator.  (We use the number~$0$ in lieu of a unit value.)}
\label{fig:self-eval}
\end{figure}
We thus map each object term~$e$ to an object term~$\encode{e}$ as follows.
We call this mapping \emph{pre-encoding}.
\begin{align*}
    \encode{x} &= x \\
    \encode{n} &= \ident{int}\, n \\
    \encode{e_1 + e_2} &= \ident{add} \encode{e_1} \encode{e_2} \\
    \encode{\cond{b}{t}{e}} &= \rlap{$\ident{if\_} \encode{b}
        \left(\fun{\_}\encode{t}\right) \left(\fun{\_}\encode{e}\right)$}
        \displaybreak[0] \\
    \encode{\fun{x}e} &= \ident{lam} (\fun{x} \encode{e}) \\
    \encode{fx} &= \ident{app} \encode{f} \encode{x} \\
    \encode{\fix{f}e} &= \ident{fix} (\fun{f} \encode{e})
\end{align*}
The metavariables $x$ and~$n$ stand for a variable and an integer,
respectively.
This pre-encoding is just like how we represent object terms in the
metalanguage in the preceding sections, but it produces
terms in the object language rather than the metalanguage.

@@

\oleg{The following are almost unedited excerpts from our
  correspondence with Sumii about Birkedal's thesis. We definitely
  should include this discussion and the reference to his thesis.}
\Citet{Birkedal-PE-ML} have hand-written a compiler generator (cogen)
for the Core Standard ML, which transforms a program into its tagless
generating extension (and therefore serves as a tagless partial
evaluator, intepreter, and compiler in our sense). They still have to
use the "universal type" at the cogen level, which is removed from
compile-time, so that the end result is tagless indeed.  The main
difference between the work of \Citet{Birkedal-PE-ML} and ours is how
\emph{early} it becomes \emph{manifest} to the metalanguage
implementation that our code is safe.  The fact that
\Citet{Birkedal-PE-ML} 's generating extension never generates a
specialized program that is ill-typed is not manifest to the SML
compiler until the SML compiler sees each particular specialized
program.  Moreover and analogously, the fact that
\Citet{Birkedal-PE-ML} 's cogen never generates a generating extension
that is ill-typed (or that may generate a specialized program that is
ill-typed) is not manifest to the SML compiler until the SML compiler
sees each particular generating extension (or each particular
specialized program).  This early and manifest type safety is the
payoff of our treating an object program not as a data structure but
as calls to abstract-syntax constructor functions.  The tradeoff, as
we currently see it, is that we have to swear off intensional analysis
of code. \oleg{Unless we convert to our representation to GADT, which
  is always possible, if GADTs are available.}
